---
title: "Benchmarks"
author: "Johan Larsson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette contains benchmarks of **sgdnet** against other similar
packages. The data has been precomputed from scripts that are
available at <https://github.com/jolars/sgdnet/data-raw/>.

The benchmarks were produces by running the packages over a range
of tolerance thresholds and recording the run times.

# Binomial family

In this section, we are going to look at the following datasets:

Name              Observations      Features
----------    ----------------     ---------
icjnn1                  49,990            22
a9a                     32,561           123
phishing                11,055            68
mushroooms               8,124           112
covtype                581,012            54

Table: Benchmarking datasets for the binomial model.

All of these have been collected from the
[libsvm binary dataset collection](https://www.csie.ntu.edu.tw/~cjlin/libsvmtools/datasets/binary.html).


```{r, fig.cap = "Benchmarking results for binomial responses."}
library(lattice)
library(latticeExtra)
library(directlabels)
p_binomial <- lattice::xyplot(mean_loss ~ time | dataset + penalty,
                              data_benchmarks_binomial,
                              type = "l",
                              xlab = "Time (normalized)",
                              ylab = "Objective loss",
                              groups = package,
                              scales = list(y = list(relation = "free")),
                              auto.key = list(lines = TRUE, points = FALSE))
p <- useOuterStrips(p_binomial)
direct.label(p, "angled.boxes")
```

